Simulating a flock of birds

### 2017-10-29 tex/article.tex

\usepackage{subfig}

\usepackage{todonotes}

\usepackage{blindtext} % Package to generate dummy text throughout this template

\usepackage[sc]{mathpazo} % Use the Palatino font
\usepackage[T1]{fontenc} % Use 8-bit encoding that has 256 glyphs
\linespread{1.05} % Line spacing - Palatino needs more space between lines
\usepackage{microtype} % Slightly tweak font spacing for aesthetics

\usepackage[english]{babel} % Language hyphenation and typographical rules

\usepackage[hmarginratio=1:1,top=32mm,columnsep=20pt]{geometry} % Document margins
\usepackage[labelfont={bf,up},textfont={it,up}]{caption} % Custom captions under/above floats in tables or figures
\usepackage{booktabs} % Horizontal rules in tables

@@

\usepackage{titling} % Customizing the title section

\usepackage{hyperref} % For hyperlinks in the PDF

\usepackage{amssymb,amsmath,bbold,braket}

@@

\pretitle{\begin{center}\Huge\bfseries} % Article title formatting
\posttitle{\end{center}} % Article title closing formatting
\title{Simulating a flock of birds} %Simulating a flock of birds
\author{%
\textsc{Steffen Randrup, Michel Smola, Andrei Timus}\\[1ex] % Your name
\normalsize University of Copenhagen \\ % Your institution
%\normalsize \href{mailto:john@smith.com}{john@smith.com} % Your email address
%\and % Uncomment if 2 authors are required, duplicate these 4 lines if more
%\textsc{Jane Smith}\thanks{Corresponding author} \\[1ex] % Second author's name
%\normalsize University of Utah \\ % Second author's institution
%\normalsize \href{mailto:jane@smith.com}{jane@smith.com} % Second author's email address
}
\date{\today} % Leave empty to omit a date
\renewcommand{\maketitlehookd}{%
\begin{abstract}
\noindent %\blindtext % Dummy abstract text - replace \blindtext with your abstract text

In this paper we present a model for studying the behaviour of a group of particles which have some level of interaction between them. The model was inspired from the~\cite{Vicsek} paper\todo[inline]{Present paper and author by name. Also it probably shouldn't be in the abstract} and as in~\cite{Vicsek} the particles have a constant absolute velocity and at each time step  assume the average direction of nearby neighbours with some added perturbation, $\eta$. The behaviour was studied by analysing how the average velocity of the flock $v_a$ varied when parameters like noise, density, number of particles where changed. In addition to $[1]$ we have introduced a cone of vision for every particle and studied the behaviour of the flock in the same conditions. From the simulation results it has been observed that the flock goes through a phase transition where all the particles assume the same direction. The average velocity $v_a$ played the role of an order parameter and it scaled with $(\eta_c-\eta)^\beta$, where $\beta$ is a critical exponent.

\end{abstract}
}

@@

\begin{document}

% Print the title
\maketitle

@@

\begin{figure}[!htb]
	\centering
	\subfloat{{\includegraphics[width=\columnwidth/2]{low-eta-low-rho}}}
  \subfloat{{\includegraphics[width=\columnwidth/2]{low-eta-high-rho}}}
	\qquad
	\subfloat{{\includegraphics[width=\columnwidth/2]{high-eta-low-rho}}}
	\subfloat{{\includegraphics[width=\columnwidth/2]{high-eta-high-rho}}}
	\caption{test for figures}
  \label{fig:configs}
\end{figure}

Figure~\ref{fig:configs} shows the configuration of birds for different values of $\rho$ and $\eta$.  
If the density is high enough, while the noise has low values, the system goes through a kinetic phase transition where all the particles share the same direction. In order to characterize the behaviour of the flock we have determined the average normalized velocity of the system and investigated how it changes when the input parameters are modified:

@@

For the disordered phase $v_a$ is aproximatively 0, while for the ordered phase $v_a$ has the value of 1.
\todo[inline]{refrace}

@@

\subsection{Finding $\eta_c$}
To determine the critical noise parameter $\eta_c$ we used a quadratic fit model
to to fit the falling curve for $\eta < \eta_c$ and a linear fit for $\eta > 
\eta_c$, both visible in Fig.~\ref{fig:va_over_eta}. $\eta_c$ is then set to 
the intersection of the fit curves.

\begin{equation}
  \label{eq:va-eta}
  v_a \approx[\eta_c(\rho)-\eta]^\beta 
\end{equation}
\begin{equation}
  \label{eq:va-rho}
  v_a \approx[\rho-\rho_c(\eta)]^\gamma
\end{equation}

We further extend the model by implementing a cone-like vision. This means, that rather than averaging the velocities over all birds within a certian radius, we only look at a slice of this circle. For a bird to be within the vision of another bird, it must be within the interaction radius, $r$, and the angle between the birds velocities must be within $\pm \theta_{cone}$. In the limit $\theta_{cone} = 180 \deg$\todo[inline]{use radians for paper?} we should see the same behavior of the model as without implementation of this vision.

@@

The alignment, $v_a$, depends on the amount of randomness, $\eta$, in the system and the density, $\rho$, as seen in equations~\eqref{eq:va-eta} and \eqref{eq:va-rho}. In Figure\ref{fig:va_over_eta} we see how, the alignment of the birds decreases as the randomness of their motion increases. The decrease is to be expected as a higher degree of randomness should disrupt alignment. What might be more interesting is the general shape of the curves. We may then determine $\beta$ from equation\eqref{eq:va-eta} to be \todo[inline]{get value of $\beta$} \todo[inline]{What do we use $\beta$ for?}. This is also shown in figure~\ref{fig:criticaleta}. The results presented here are all computed without the conevision (or with $\theta_{cone} = 180\deg$)\todo[inline]{use radians for paper?}. The effects of it will be presented in \ref{subsec:results_conevision}.

\todo[inline]{
TO EDIT
For the disordered phase $v_a$ is aproximatively 0, while for the ordered phase $v_a$ has the value of 1.
%rephrase:
In the following figure we show how $v_a$ varies with the noise $\eta$ and density  $\rho$ for a constant value of density and noise respectively.

As it can be seen from the graph the overall alignment of the particles depends on the noise and for a high number of particles N the system becomes more disordered.

For high values of density the flock reaches a coherent moving phase, while for low values of density the particles move randomly in the lattice.

TO EDIT

}

\begin{figure}[!htb]
  \centering
  \includegraphics[width=\columnwidth]{va_over_eta}
  \caption{Dependance of alignment on randomness}
  \label{fig:va_over_eta}
\end{figure}

\begin{figure}[!htb]
	\centering
	\includegraphics[width=\columnwidth]{va_over_etac_minus_eta_over_etac}
	\caption{Determining $\beta$}
  \label{fig:criticaleta}
\end{figure}

@@

\begin{figure}[!htb]
	\centering
	\includegraphics[width=\columnwidth]{va_over_rho}
	\caption{Dependance of alignment on density}
  \label{fig:va_over_rho}
\end{figure}

@@

The effects of our implementation of an angle based vision are only detectable on both low randomness and low density. This is expected as a higher density causes an effective long range interaction, which may span the entire flock\todo[inline]{Is long range interaction described?}. This negates the effect of an angle based vision as the flock moves in a coherent fashion. As for randomness of motion it decreases the overall alignment of the flock as seen in figure~\ref{fig:va_over_eta}. In figure~\ref{fig:conevision} we present the overall alignment's dependance of the angle describing this cone-like vision for different densities. The randomness is kept at a low value of $\eta=0.1$. The results correspond quite well with our expectation of increasing the alignment with larger angle and for an angle of 180 degrees we see, the alignment looks like it did when we considered everything within a circle. We also notice the effects of the density of birds. As described above we expect a long range interaction between the birds to arise when the density increases. This effect is visible in the figure as we see the alignment increase faster, when the density is higher.

\begin{figure}[!htb]
  \begin{center}
    \includegraphics[width=\columnwidth]{va_over_angle}
  \end{center}
  \caption{Conevision}
  \label{fig:conevision}
\end{figure}

\todo[inline]{There is something about how the alignment only starts increasing once the angle increases above 25 degrees}

@@

We have compared the collective motion of birds to a thermodynamic phase transition. The connection between these is not obvious. However we see simlilarities in how increasing the randomness\todo[inline]{temperature?} will distrupt alignment and how denser systems will display longer effective interaction range.

The addition of a cone-like vision is, as the name suggests, inspired from birds rather than thermodynamics. To our knowledge it doesn't have an equivalent in this field\todo[inline]{does it?}, \todo[inline]{however it might tell us something about directional interaction}

@@

In this paper we have presetend a model for studying the behaviour of interacting particles. The interaction between particles consisted in maintaining the average velocity of nearby neighbours. The system was studied when particles shared a common interaction radius and when each particle had a cone of vision. During the simulation it has been observed that the system goes through a phase transition where all particles assume the same direction. In order to describe the transition we have calculated the average normalized velocity va and investigated how it behaves when we modify the parameters of the system. In the thermodynamic limit va can be described using (5) or (6).
By plotting these equations we have found  the critical exponent $\beta$ to be equal to ....

>>>>

\usepackage{subfig}
\usepackage{todonotes}
\usepackage[sc]{mathpazo} % Use the Palatino font
\usepackage[T1]{fontenc} % Use 8-bit encoding that has 256 glyphs
\linespread{1.05} % Line spacing - Palatino needs more space between lines
\usepackage{microtype} % Slightly tweak font spacing for aesthetics
\usepackage[english]{babel} % Language hyphenation and typographical rules
\usepackage[hmarginratio=1:1,top=32mm,columnsep=20pt]{geometry} % Document margins
\usepackage[labelfont={bf,up},textfont={it,up}]{caption} % Custom captions under/above floats in tables or figures
\usepackage{booktabs} % Horizontal rules in tables

@@

\usepackage{titling} % Customizing the title section
\usepackage{hyperref} % For hyperlinks in the PDF
\usepackage{amssymb,amsmath,bbold,braket}

@@

\pretitle{\begin{center}\Huge\bfseries} % Article title formatting
\posttitle{\end{center}} % Article title closing formatting
\title{Simulating a flock of birds} %Simulating a flock of birds
\author{%
\textsc{Steffen Randrup, Michel Smola, Andrei Timus}\\[1ex] % Your name
\normalsize University of Copenhagen \\ % Your institution
%\normalsize \href{mailto:john@smith.com}{john@smith.com} % Your email address
%\and % Uncomment if 2 authors are required, duplicate these 4 lines if more
%\textsc{Jane Smith}\thanks{Corresponding author} \\[1ex] % Second author's name
%\normalsize University of Utah \\ % Second author's institution
%\normalsize \href{mailto:jane@smith.com}{jane@smith.com} % Second author's email address
}
\date{\today} % Leave empty to omit a date
\renewcommand{\maketitlehookd}{%
\begin{abstract}
\noindent
In this paper we present a model for studying the behaviour of a group of particles which have some level of interaction between them. The model was inspired from the~\cite{Vicsek} paper\todo[inline]{Present paper and author by name. Also it probably shouldn't be in the abstract} and as in~\cite{Vicsek} the particles have a constant absolute velocity and at each time step  assume the average direction of nearby neighbours with some added perturbation, $\eta$. The behaviour was studied by analysing how the average velocity of the flock $v_a$ varied when parameters like noise, density, number of particles where changed. In addition to $[1]$ we have introduced a cone of vision for every particle and studied the behaviour of the flock in the same conditions. From the simulation results it has been observed that the flock goes through a phase transition where all the particles assume the same direction. The average velocity $v_a$ played the role of an order parameter and it scaled with $(\eta_c-\eta)^\beta$, where $\beta$ is a critical exponent.
\end{abstract}
}

@@

\begin{document}
\maketitle

@@

\begin{figure}[!htb]
	\centering
	\subfloat{{\includegraphics[width=\columnwidth/2]{low-eta-low-rho}}}
  \subfloat{{\includegraphics[width=\columnwidth/2]{low-eta-high-rho}}}
	\qquad
	\subfloat{{\includegraphics[width=\columnwidth/2]{high-eta-low-rho}}}
	\subfloat{{\includegraphics[width=\columnwidth/2]{high-eta-high-rho}}}
	\caption{Different configurations of $\eta$ and $\rho$\\Top left: low $\eta$ low $\rho$\\Top right: Low $\eta$, high $\rho$\\Bottm left: High $\eta$, low $\rho$\\Bottom right: High $\eta$, high $\rho$}\label{fig:configs}
\end{figure}

Figure~\ref{fig:configs} shows the configuration of birds for different values of $\rho$ and $\eta$.  
For low values of both $\rho$ and $\eta$ the birds tend to form smaller groups in which the birds move in the same direction.
The groups themselves move randomly.
When the randomness is high and the density low, the overall motion of the birds is just random.
In the case of high randomness and high density, the birds appear to move randomly but with an overall trend in directions
shared among the all the birds.
If the density is high enough, while the noise has low values, the virds move ordly in one direction.
This overall trend in motion is similar to the system going through a kinetic phase transition
where all the particles share the same direction. In order to characterize the behaviour of the flock
we have determined the average normalized velocity of the system and investigated how it changes when the input parameters are modified:

@@

For disordered motion $v_a$ is aproximatively 0, while for ordered $v_a$ has the value near 1.

@@

\subsection{Finding $\eta_c$}
To determine the critical noise parameter $\eta_c$ we used a quadratic fit model
to to fit the falling curve for $\eta < \eta_c$ and a linear fit for $\eta > 
\eta_c$, both visible in figure~\ref{fig:va_over_eta}. $\eta_c$ is then set to 
the intersection of the fit curves. Hereby we determine
\begin{align}
\eta_c = 4.56***
\end{align}
\todo[inline]{put in final number}

\begin{equation}
  \label{eq:va-eta}
  v_a \approx{[\eta_c(\rho)-\eta]}^\beta 
\end{equation}
\begin{equation}
  \label{eq:va-rho}
  v_a \approx{\left[\rho-\rho_c (\eta)\right]}^\gamma
\end{equation}

We further extend the model by implementing a cone-like vision. This means, that 
rather than averaging the velocities over all birds within a certian radius, we 
only look at a slice of this circle. For a bird to be within the vision of 
another bird, it must be within the interaction radius, $r$, and the angle 
between the birds velocities must be within $\pm \theta_{\text{cone}}$. The 
limit $\theta_{\text{cone}} = 2\pi$ equals the case without limiting the
cone vision and reprocuces the results from~\cite{Vicsek}.

@@

The alignment, $v_a$, depends on the amount of randomness $\eta$ in the system 
and the density $\rho$, as seen in equations~\eqref{eq:va-eta} and~\eqref{eq:va-rho}.
In figure~\ref{fig:va_over_eta} we see how the alignment of 
the birds decreases as the randomness of their motion increases. The decrease 
is to be expected, as a higher degree of randomness should disrupt alignment. 
This reflects to $v_a$, which approaches $0$ in a totally unordered state and $1$
in a totally ordered state.
What might be more interesting is the general shape of the curves. We may then 
determine $\beta$ from equation~\eqref{eq:va-eta} to be \todo[inline]{get value of $\beta$}
This is also shown in figure~\ref{fig:criticaleta}. The results presented here 
are all computed without the conevision (equivalent to $\theta_{\text{cone}} = 2\pi$). 
Figure~\ref{fig:va_over_eta} shows how $v_a$ varies with the noise $\eta$ and 
density $\rho$ for a constant value of density and noise respectively.
As seen in the graph the overall alignment of the particles depends 
on the noise and for a high number of particles N the system becomes less ordered. 
For high $\eta$ the flock reaches a coherent moving phase, while for low $\eta$ 
the particles move randomly in the lattice.

\begin{figure}[!htb]
  \centering
  \includegraphics[width=\columnwidth]{va_over_eta}
  \caption{Dependence of alignment on randomness}\label{fig:va_over_eta}
\end{figure}

\begin{figure}[!htb]
	\centering
	\includegraphics[width=\columnwidth]{va_over_etac_minus_eta_over_etac}
	\caption{Determining $\beta$}\label{fig:criticaleta}
\end{figure}

@@

\begin{figure}[!htb]
	\centering
	\includegraphics[width=\columnwidth]{va_over_rho}
	\caption{Dependance of alignment on density}\label{fig:va_over_rho}
\end{figure}

@@

The effects of our implementation of an angle based vision are only detectable on both low randomness and low density.
This is expected as a higher density causes an effective long range interaction, which may span the entire flock.
This negates the effect of an angle based vision as the flock moves in a coherent fashion.
As for randomness of motion it decreases the overall alignment of the flock as seen in figure~\ref{fig:va_over_eta}.
In figure~\ref{fig:conevision} we present the overall alignment's dependance of the angle describing this cone-like
vision for different densities. The randomness is kept at a low value of $\eta=0.1$.
The results correspond quite well with our expectation of increasing the alignment with larger angle.
For an angle of 180 degrees we see, the alignment looks like it did when we considered every bird within the interaction radius.
We also notice the effects of the density of birds. As described above we expect a long range interaction
between the birds to arise when the density increases.
This effect is visible in the figure as we see the alignment increase faster, when the density is higher.

\begin{figure}[!htb]
\begin{center}
\includegraphics[width=\columnwidth]{va_over_angle}
\end{center}
\caption{Conevision}\label{fig:conevision}
\end{figure}

@@

We have compared the collective motion of birds to a phase transition.
The connection between these is not obvious. However we do see simlilarities in how increasing the noise in the system
will distrupt alignment and how denser systems will display longer effective interaction range. Causing the entire
system to behave coherently.

The addition of a cone-like vision is, as the name might suggests, inspired from birds rather than the study of phase transitions.
The effects of the vision based approach were suttle. They were only detectable for low densities and noise, and for
sufficiently large angles of vision the system would again display behavior to that without the vision based approach.

@@

In this paper we have presetend a model for studying the behaviour of interacting particles. The interaction between particles consisted in maintaining the average velocity of nearby neighbours. The system was studied when particles shared a common interaction radius and when each particle had a cone of vision. During the simulation it has been observed that the system goes through a phase transition where all particles assume the same direction. In order to describe the transition we have calculated the average normalized velocity va and investigated how it behaves when we modify the parameters of the system. In the thermodynamic limit va can be described using (5) or (6).
By plotting these equations we have found  the critical exponent $\beta$ to be equal to\todo[inline]{$\beta$}
